The forestecology R package for fitting and assessing neighborhood models of the effect of interspecific competition on the growth of trees

### 2021-03-09 paper/forestecology.Rmd
4210 -> 4069

## Step 1: Compute the growth of trees based on census data {compute-growth}

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. The `compute_growth()` function computes average annual growth assuming census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling and checking. For example, the SCBI site records all diameters at breast height (DBH) in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site records them in centimeters [@allen_michigan_2020].

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is to check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.] Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then supplied as arguments to the `compute_growth()` function, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`), since we are only interested in the diameter growth of surviving, and not resprouted, stems.

@@

Site data that does not align with this convention will need to be transformed for use with the `compute_growth()` function. However, in the end, all that matters is that the growth of all stems is saved in a data frame of class `sf` and, at a minimum, contains the variable uniquely identifying each stem, `sp`, `dbh1`, `growth`, `geometry`.

@@

The next step in our analysis sequence is to add additional spatial information to our main `growth_scbi` data frame. The first element of spatial information we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative (such as competitor basal area), we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of study regions, for which all neighbors will not be included in the censused stems. In order to account for such edge effects, only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growth modeled [@waller_applied_2004].

Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest: all trees within a distance `comp_dist` of a focal tree are considered its competitors (assuming the same units as the `gx` and `gy` location variables). In our case we set this value at 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. Using this value along with a manually constructed `sf` object representation of the study region's boundary via its vertices, we apply the `add_buffer_variable()` to our `growth_scbi` data frame to add a `buffer` boolean variable: all trees who have `buffer` set to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only be considered as competitor trees whose growth will not.

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, many of these algorithms assume that the observations are independent of each other. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees [@roberts_cross-validation_2017 @pohjankukka_estimating_2017]
In the example below, we first manually define four folds that partition the study region as an `sf` object. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct fold (saved in the `foldID` variable) [@valavi_blockcv_2019]. ^[In the Supporting Information we present an example where the folds themselves are also created using the `spatialBlock()` function given a specified `cv_block_size`.]

@@

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of adding these two elements of information to the `growth_scbi` data frame. The location of each tree is marked with an integer indicating which fold it belongs to, where the folds are marked with solid lines. The color of each digit indicates whether the tree is part of the buffer region (and thus will only be considered as a competitor tree in our model) or is part of the interior of the study region (and thus is a focal tree whose growth is of modeled interest).

@@

The next step in our analysis sequence is to identify all focal trees and their corresponding competitor trees. More specifically, it identifies all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist` of it.  The `create_focal_vs_comp()` functions performs these tasks and returns a new data frame of type `sf`. On top of the previous arguments `comp_dist` defining the competition neighborhood and `id` indicating which variable uniquely identifies each tree-stem, this function also requires an `sf` object representation of the spatial cross-validation blocks/folds. In this example, the blocks were manually encoded in `blocks_scbi` by specifying it's vertices in Section \ref{spatial-information}^[We present an alternative method for defining spatial cross-validation blocks is using the `spatialBlock()` function from the `blockCV` package in the Supporting Information.]. We present the resulting data frame below with the `foldID` variable omitted for compactness of presentation.

@@

The resulting data frame `focal_vs_comp_scbi` has `r nrow(focal_vs_comp_scbi)` rows, representing the subset of the `r nrow(growth_scbi)` trees in `growth_scbi` that will be considered as focal trees. Two new variables `focal_ID` and `focal_sp` relate to tree-stem identification and species information. Most notably however is a new variable `comp` which contains information on all competitor trees for a given focal tree, saved in `tidyr` package list-column format [@tidyr_package]. For example, we drill-down on the tree with `focal_ID` 4, which has 20 competitor trees each described by 4 variables as indicated by the fact that `comp` is a `tibble [20 × 4]`.

@@

Using the `unnest()` function from the `tidyr` package, we can flatten list-column into regular columns. We observe that for the same focal tree, we have information on all 20 competitor trees whose `dist` distance to the focal tree is $\leq$ `r comp_dist`: their unique tree-stem ID number, their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ where $DBH$ is the value from the earlier of the two censuses in cm. Saving our focal versus competitor information in list-column minimizes redundancy since we do not repeat information on the focal tree 20 times.

@@

Now that we've identified all focal and corresponding competitor trees and saved this information in a data frame of type `focal_vs_comp`, the final step in our analysis sequence is to fit a model for the growth of all focal trees. Currently the `forestecology` package can only fit the competition Bayesian linear regression model outlined in Section \ref{competition-model} using the `comp_bayes_lm()` function. However, any model implemented in a function that similarly takes an input data frame of type `focal_vs_comp` as an argument can also be used. For our specific competition Bayesian linear regression model, we also specify prior distributions on all parameters of interest (here chosen to be the defaults as specified in `?comp_bayes_lm`).

@@

In order to evaluate the effect of competitor species identity, we use the four steps of our analysis sequence answer along with a permutation test: Under a null hypothesis where competitor species identity does not matter, we can permute/shuffle this variable within each focal tree, compute the RMSE (the test statistic of interest), repeat this process several times to construct a null distribution of the RMSE, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test randomly resamples the `comp_sp` variable with replacement, leaving all other variables intact. The resampling with replacement is nested within each focal tree in order to preserve the neighborhood structure of our competition model. To run the permutation test, we use the same `comp_bayes_lm()` function as in Section \ref{model-fit-predict}, but with a `run_shuffle = TRUE` argument.

@@

This algorithm is implemented in the `run_cv()` function, which is a wrapper function to the `comp_bayes_lm()` function that fits the model and the `predict()` generic that returns fitted/predicted values. We compare these values to the observed growth values to again compute our RMSE.

@@

The `run_cv` function also accepts the `run_shuffle` argument. This permutes the competitor species, as described above, but does so when calculating predicted growth with the cross validated scheme. Figure \ref{fig:scbi-simulation} compares model performance when permuting competitor species and calculating RMSE with and without cross validation. Without cross-validation the competitor identity did matter, the non-permuted competitor species had a much lower RMSE than the permuted one. But once we include the spatial cross validation this improvement disappears. So in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other places (@allen_permutation_2020 @uriarte_spatially_2004). This highlights the importance of cross validation, without it the model was overfit.

>>>>

## Step 1: Compute the growth of trees based on census data {#compute-growth}

The first step in the our analysis sequence is to compute the growth of trees using data from two censuses. `compute_growth()` computes average annual growth assuming census data that roughly follows ForestGEO standards. Despite such standards, minor variations will still exist between sites, thereby necessitating some data wrangling and checking. For example, the SCBI site records all diameters at breast height (DBH) in millimeters [@bourg_initial_2013], whereas the Michigan Big Woods site records them in centimeters [@allen_michigan_2020].

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020 [@gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020]. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters. Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

@@

These two data frames are then supplied as arguments to `compute_growth()`, along with the `id` argument that specifies the variable that uniquely identifies each tree-stem. Note furthermore that we discard all resprouts in the later census (those with `code == R`), since we are only interested in the diameter growth of surviving, and not resprouted, stems.

@@

Site data that does not align with this convention will need to be transformed for use with `compute_growth()`. However, in the end, all that matters is that the growth of all stems is saved in a data frame of class `sf` and, at a minimum, contains the variable uniquely identifying each stem, `sp`, `dbh1`, `growth`, `geometry`.

@@

The next step in our analysis sequence is to add additional spatial information to `growth_scbi`. The first element of spatial information we add is a "buffer region" to the periphery of the study region. Since some of our model's explanatory variables are cumulative (such as competitor basal area), we must ensure that all trees being modeled are not biased to have different neighbor structures. This is of concern for trees at the boundary of study regions, for which all neighbors will not be included in the censused stems. In order to account for such edge effects, only trees that are not part of this buffer region, i.e. are part of the interior of the study region, will have their growth modeled [@waller_applied_2004].

Our model of interspecific competition relies on a spatial definition of who the competitor trees are for focal trees of interest: all trees within a distance `comp_dist` of a focal tree are considered its competitors (assuming the same units as the `gx` and `gy` location variables). In our case we set this value at 7.5m, a value informed by other studies [@canham_neighborhood_2004, @uriarte_spatially_2004, @canham_neighborhood_2006]. Using this value along with a manually constructed `sf` object representation of the study region's boundary via its vertices, we apply the `add_buffer_variable()` to `growth_scbi` to add a `buffer` boolean variable: all trees who have `buffer` set to `FALSE` will be our focal trees whose growth will be modeled, whereas those with `TRUE` will only be considered as competitor trees whose growth will not.

@@

The second element of spatial information are blocks corresponding to folds of a spatial cross-validation algorithm used to estimate out-of-sample model error. Conventional cross-validation algorithms assign observations to folds by randomly resampling individual observations. However, many of these algorithms assume that the observations are independent of each other. In the case of forest census data, observations exhibit spatial autocorrelation. We therefore incorporate this spatial dependence into the cross-validation algorithm with our spatial blocks of trees [@roberts_cross-validation_2017 @pohjankukka_estimating_2017]
In the example below, we first manually define four folds that partition the study region as an `sf` object. We then use the output of the `spatialBlock()` function from the `blockCV` package to associate each tree in `growth_scbi` to the correct fold (saved in the `foldID` variable) [@valavi_blockcv_2019]. ^[In the Supporting Information we present an example where the folds themselves are also created using `spatialBlock()` given a specified `cv_block_size`.]

@@

Figure \ref{fig:scbi-spatial-information} illustrates the net effect of adding these two elements of information to `growth_scbi`. The location of each tree is marked with an integer indicating which fold it belongs to, where the folds are marked with solid lines. The color of each digit indicates whether the tree is part of the buffer region (and thus will only be considered as a competitor tree in our model) or is part of the interior of the study region (and thus is a focal tree whose growth is of modeled interest).

@@

The next step in our analysis sequence is to identify all focal trees and their corresponding competitor trees. More specifically, it identifies all trees that are not part of the buffer region, have a valid `growth` measurement, and have at least one neighbor within `r comp_dist` of it.  The `create_focal_vs_comp()` functions performs these tasks and returns a new data frame of type `sf`. On top of the previous arguments `comp_dist` defining the competition neighborhood and `id` indicating which variable uniquely identifies each tree-stem, this function also requires an `sf` object representation of the spatial cross-validation blocks/folds. In this example, the blocks were manually encoded in `blocks_scbi` by specifying it's vertices in Section \ref{spatial-information}. We present the resulting data frame below with the `foldID` variable omitted for compactness of presentation.

@@

The resulting `focal_vs_comp_scbi` has `r nrow(focal_vs_comp_scbi)` rows, representing the subset of the `r nrow(growth_scbi)` trees in `growth_scbi` that will be considered as focal trees. Two new variables `focal_ID` and `focal_sp` relate to tree-stem identification and species information. Most notably however is a new variable `comp` which contains information on all competitor trees for a given focal tree, saved in `tidyr` package list-column format [@tidyr_package]. For example, we drill-down on the tree with `focal_ID` 4, which has 20 competitor trees each described by 4 variables as indicated by the fact that `comp` is a `tibble [20 × 4]`.

@@

Using `unnest()` from the `tidyr` package, we can flatten list-column into regular columns. We observe that for the same focal tree, we have information on all 20 competitor trees whose `dist` distance to the focal tree is $\leq$ `r comp_dist`: their unique tree-stem ID number, their species, and their basal area (in m$^2$) calculated as $\frac{\pi \times (\text{DBH/2})^2}{10000}$ where $DBH$ is the value from the earlier of the two censuses in cm. Saving our focal versus competitor information in list-column minimizes redundancy since we do not repeat information on the focal tree 20 times.

@@

The final step is to fit a model for the growth of all focal trees. Currently the `forestecology` package can only fit the competition Bayesian linear regression model outlined in Section \ref{competition-model} using `comp_bayes_lm()`. However, any model implemented in a function that similarly takes an input data frame similar to `focal_vs_comp` as an argument can also be used. For our specific competition Bayesian linear regression model, we also specify prior distributions on all parameters of interest (here chosen to be the defaults as specified in `?comp_bayes_lm`).

@@

In order to evaluate the effect of competitor species identity, we use the four steps of our analysis sequence answer along with a permutation test: Under a null hypothesis where competitor species identity does not matter, we can permute/shuffle this variable within each focal tree, compute the RMSE (the test statistic of interest), repeat this process several times to construct a null distribution of the RMSE, and compare it to the observed RMSE to assess significance. Going back to our example in Section \ref{focal-vs-comp} of focal tree with `focal_ID` 4 and its 20 competitors, the permutation test randomly resamples the `comp_sp` variable with replacement, leaving all other variables intact. The resampling with replacement is nested within each focal tree in order to preserve the neighborhood structure of our competition model. To run the permutation test, we use `comp_bayes_lm()` as in Section \ref{model-fit-predict}, but with a `run_shuffle = TRUE` argument.

@@

This algorithm is implemented in `run_cv()`, which is a wrapper function to `comp_bayes_lm()` that fits the model and the `predict()` generic that returns fitted/predicted values. We compare these values to the observed growth values to again compute our RMSE.

@@

`run_cv()` also accepts the `run_shuffle` argument. This permutes the competitor species, as described above, but does so when calculating predicted growth with the cross validated scheme. Figure \ref{fig:scbi-simulation} compares model performance when permuting competitor species and calculating RMSE with and without cross validation. Without cross-validation the competitor identity did matter, the non-permuted competitor species had a much lower RMSE than the permuted one. But once we include the spatial cross validation this improvement disappears. So in this 9 ha subplot of the SCBI plot competitive interactions do not depend on the identity of the competitor, which is the opposite of what has been observed in other places (@allen_permutation_2020 @uriarte_spatially_2004). This highlights the importance of cross validation, without it the model was overfit.
